---
title: "Code for 'Heavy metals in the Arctic: Distribution and enrichment of five metals in Alaskan soils'"
author: "Clarice Perryman, Jochen Wirsing, Apryl Perry""
date: "February 28, 2019"
output: html_document
---
The following code conducts the data filtering, analysis, and visualization described in Perryman et al. (2020), "Heavy metals in the Arctic: Distribution and enrichment of five metals in Alaskan soils." This code was last updated 5/11/20. Please email Clarice Perryman at crp1006@wildcats.unh.edu with any questions. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Clear workspace (if desired/needed)
```{r}
rm(list = ls())
```

Load needed libraries
```{r}
library(tidyverse) #version 1.2.1
library(dplyr) #version 0.7.8
library(RCurl) #version 1.95.4.11
library(ggplot2) #version 3.1.0
library(reshape2) #version 1.4.3
library(agricolae) #this package runs the Kruskal-Wallis and post-hoc

#gets version numbers for packages installed in namespace, can reference to above version info for each package for reproducibility
print(paste("R", getRversion()))
print("-------------")
for (package_name in sort(loadedNamespaces())) {
    print(paste(package_name, packageVersion(package_name)))
}
```

Read in raw data file from Git, ignored first 4 rows (USGS logo header) - blank and or random file info, blank cells to NA
```{r}
downloadBSoil <- getURL("https://raw.githubusercontent.com/jagilman/Arctic-Ecol-Soc-2019/master/BestValue_Soil.csv") 
B_Soil <- read.csv(text = downloadBSoil, header = TRUE, skip = 4, na.strings = c(NA, ""))
```

Filtering down to desired metals and metedata. Removed all columns ending with Sum, these are repeat info of existing columns. Create new df with only desired info. 
```{r}
NoSum <- select (B_Soil, -c(ends_with('_Sum')))

USGS1 <- select(NoSum, matches('LAB_ID|FIELD_ID|DATE_SUBMITTED|DATE_COLLECT|LATITUDE|LONGITUDE|QUAD|SPHEROID|DATUM|DEPTH|SAMPLE_SOURCE|LOCATE_DESC|PRIMARY_CLASS|SAMPLE_COMMENT|SAMPLE_ZONE|HORIZON|ORGANICS|DRAINAGE|PREP|Hg_|Pb_|Ni_|As_|Cr_|Tl_'))


```

Filtering everything that has no info on the following:

* depth
* sample source

Then narrowed down columns futher to final metals of interest.  

```{r latlon_depth}

USGS2 <- filter(USGS1, is.na(DEPTH) == FALSE)
USGS3 <- filter(USGS2, is.na(SAMPLE_SOURCE) == FALSE)
USGS4 <- select(USGS3,FIELD_ID, DATE_COLLECT, DEPTH, LOCATE_DESC, HORIZON, DATUM, LATITUDE, LONGITUDE, SAMPLE_SOURCE, As_ppm, As_ppm_AM, Cr_ppm, Cr_ppm_AM, Hg_ppm, Hg_ppm_AM, Ni_ppm, Ni_ppm_AM, Pb_ppm, Pb_ppm_AM) #dropped Tl - no data.
```

Cleaning data to get depth data in consistent formats and writing out file to manually convert units and code depth in Excel.All units converted to cm in Excel. Depth bins coded in excel: 0-10cm = surface, 10-30cm = mid, > 30cm = deep. 
```{r}
#Recoding a chunk of depths to separate them correctly, removing non-numeric depths
USGS4$DEPTH <- recode(USGS4$DEPTH, "surface (0-25cm)"  = "0-25 cm")
USGS4$DEPTH <- recode(USGS4$DEPTH, "0- 5cm"  = "0-5 cm")#renaming several rows to get depth intervial
USGS4 <- filter(USGS4,!DEPTH  %in% c("unknown", "by horizon", "various", "A horizon", "B horizon", "C horizon", "surface", "Various", "loess", "loess ?")) 
##making units more consistent
USGS4$DEPTH<-str_replace(USGS4$DEPTH, "inches", "in")
USGS4$DEPTH<-str_replace(USGS4$DEPTH, "feet", "ft")
USGS4$DEPTH<-str_replace(USGS4$DEPTH, "'", " in")

#separate out units
USGS5 <- separate(USGS4, col="DEPTH", c("DEPTH", "UNIT"), sep = ' ')
#separate upper and lower units
USGS5 <- separate(USGS5, col="DEPTH", c("UPPPER", "LOWER"), sep = '-') %>% mutate_at(vars("UPPPER", "LOWER"), parse_number)

#remove bad split
USGS5 <- filter(USGS5, UNIT !="channel" )
#keep only rows with units
USGS5 <- filter(USGS5, UNIT %in% c("cm", "in", "ft"))

write.csv(USGS5, file = "filtered_USGS.csv")

```

Read in CSV post coding depths, remove values < 0, recode land cover types, and order the depth factor levels for plotting. 
```{r}
filtered <- getURL("https://raw.githubusercontent.com/jagilman/Arctic-Ecol-Soc-2019/master/S1_File.csv")
USGS_analyze <- read.csv(text = filtered, header = TRUE)

#taking out values < 0
is.na(USGS_analyze$As_ppm) <- USGS_analyze$As_ppm < 0
is.na(USGS_analyze$Cr_ppm) <- USGS_analyze$Cr_ppm < 0
is.na(USGS_analyze$Hg_ppm) <- USGS_analyze$Hg_ppm < 0
is.na(USGS_analyze$Ni_ppm) <- USGS_analyze$Ni_ppm < 0
is.na(USGS_analyze$Pb_ppm) <- USGS_analyze$Pb_ppm < 0

#Renaming SAMPLE_SOURCE as "Source" for better lengend title below
colnames(USGS_analyze)[colnames(USGS_analyze)=="SAMPLE_SOURCE"] <- "Source"

#Lumping samples sources together into broader categories
USGS_analyze$Source <- recode(USGS_analyze$Source, "mine tailings"  = "mine-impacted")
USGS_analyze$Source <- recode(USGS_analyze$Source, "mine dump/prospect pit"  = "mine-impacted")
USGS_analyze$Source <- recode(USGS_analyze$Source, "forest/scattered timberland"  = "forested")
USGS_analyze$Source <- recode(USGS_analyze$Source, "permafrost"  = "tundra")
USGS_analyze$Source <- recode(USGS_analyze$Source, "shrub covered"  = "tundra")
USGS_analyze$Source <- recode(USGS_analyze$Source, "grassland/grazing land/cultivated land"  = "tundra")
USGS_analyze$Source <- recode(USGS_analyze$Source, "tundra/alpine tundra"  = "tundra")
USGS_analyze$Source <- recode(USGS_analyze$Source, "float"  = "glacial deposit")
USGS_analyze$Source <- recode(USGS_analyze$Source, "glacial debris/deposit"  = "glacial deposit")
USGS_analyze$Source <- recode(USGS_analyze$Source, "wetland/swamp/marsh/bog/fen"  = "wetland")
USGS_analyze$Source <- recode(USGS_analyze$Source, "surface"  = "wetland")

levels(USGS_analyze$Source)

#This tells R to plot the depths in this order, not alphbetically. 
USGS_analyze$depth <- factor(USGS_analyze$Depth, levels = c("surface", "mid", "deep"))
```

Looking at distributions of metals data with boxplots, Cleveland dot plots, and histograms to look at distrubtions. 
```{r}
boxplot(USGS_analyze$As_ppm)
dotchart(USGS_analyze$As_ppm, xlab = "As (ppm)", ylab = "Order of the data")
hist(USGS_analyze$As_ppm)
boxplot(USGS_analyze$As_ppm ~ USGS_analyze$Source)
USGS_analyze$As_ppm<- na_if(USGS_analyze$As_ppm, 92000.0) 
#value of 92000.0 ppm is order of mag higher than others

boxplot(USGS_analyze$Cr_ppm)
dotchart(USGS_analyze$Cr_ppm, xlab = "Cr (ppm)",  ylab = "Order of the data")
hist(USGS_analyze$Cr_ppm)
boxplot(USGS_analyze$Cr_ppm ~ USGS_analyze$Source)

boxplot(USGS_analyze$Hg_ppm)
dotchart(USGS_analyze$Hg_ppm, xlab = "Hg (ppm)", ylab = "Order of the data")
hist(USGS_analyze$Hg_ppm) 
boxplot(USGS_analyze$Hg_ppm ~ USGS_analyze$Source)
USGS_analyze$Hg_ppm<- na_if(USGS_analyze$Hg_ppm, 46400.00) 
#value of 46400.00 ppm is order of mag higher than others

boxplot(USGS_analyze$Ni_ppm)
dotchart(USGS_analyze$Ni_ppm, xlab = "Ni (ppm)",ylab = "Order of the data")
hist(USGS_analyze$Ni_ppm)
boxplot(USGS_analyze$Ni_ppm ~ USGS_analyze$Source)

boxplot(USGS_analyze$Pb_ppm)
dotchart(USGS_analyze$Pb_ppm, xlab = "Pb (ppm)", ylab = "Order of the data") 
hist(USGS_analyze$Pb_ppm)
boxplot(USGS_analyze$Pb_ppm ~ USGS_analyze$Source)
USGS_analyze$Pb_ppm<- na_if(USGS_analyze$Pb_ppm, 3800.0) 
#value of 3800.0 ppm is order of mag higher than others

#log (natural log) tranforming all metals data for display
USGS_analyze$As_log <- log(USGS_analyze$As_ppm)
USGS_analyze$Cr_log <- log(USGS_analyze$Cr_ppm)
USGS_analyze$Hg_log <- log(USGS_analyze$Hg_ppm)
USGS_analyze$Ni_log <- log(USGS_analyze$Ni_ppm)
USGS_analyze$Pb_log <- log(USGS_analyze$Pb_ppm)
```


Each of the following code chunks completes these steps for each metal:
-report summary statistics
-check normality/homoscedasticity
-natural log transform for plotting
-Kruskal-Wallis + post-hoc test of pairwise comparisons comparing land cover types and depths
-generate boxplot of concentration by land cover type and depth

These boxplots are included in Figure 2 (depth plots) and Fig S1 (land cover comparisons)

Arsenic:
```{r}
#get summary statistics for As concentrations
summary(USGS_analyze$As_ppm, na.rm=TRUE)
   #Min.      1st Qu.  Median    Mean   3rd Qu.     Max.    NA's 
    #0.39     7.00    11.00   187.83    20.00   14900.00      190 
sd(USGS_analyze$As_ppm, na.rm=TRUE)
#stavard deviation is 1120.736

#Checking distribution for stats
shapiro.test(USGS_analyze$As_ppm) #tests for normality, p < 0.001
bartlett.test(As_ppm~Source, data=USGS_analyze) #tests for homogeneity of variance, p < 0.001

#Does log transforming help meet assumptions for ANOVA?
USGS_analyze$As_log <- log(USGS_analyze$As_ppm)
shapiro.test(USGS_analyze$As_log) #no, p < 0.001
bartlett.test(As_log~Source, data=USGS_analyze)#no, p  < 0.001

#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality - this has a built in post hoc test using the criterium Fisher's least significant difference
As_bySource <- kruskal(USGS_analyze$As_ppm, USGS_analyze$Source, group=TRUE, p.adj = "bonferroni")
As_bySource #Chi-sq = 99.76, Df = 4, p < 0.001
#mine-impacted                       942.4444      a
#forested                            516.0399      b
#tundra                              497.6323      b
#glacial deposit                     496.4578      b
#sedge/moss/shrub wetland            339.5202      c

#Counts for annotation 
As_count <- select(USGS_analyze, Source, upper.cm, Depth, As_ppm)
As_count<- filter(As_count, !is.na(As_ppm))
As_count %>% 
  group_by(Source) %>%
  summarise(no_rows = length(Source))


##Arsenic boxplot
As_plot <- ggplot(USGS_analyze, aes(Source, As_log, fill=Source)) + 
   geom_boxplot() +
  xlab("source") +
  ylab("ln [As (mg/kg)]") +
  theme_classic()+
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#adds coded colors, number of obsverations, and post-hoc groups
As_plot +
  scale_fill_manual(values=c("#80b1d3", "#b3de69", "#fdb462", "#fb8072", "#bebada")) + 
  annotate("text", x = 1:5, y = -0.5, label = c("83", "188", "499", "18", "173")) +
  annotate("text", x = 1:5, y = 10, label = c("b", "b", "b", "a", "c"))

#Checking distribution of depths for stats
bartlett.test(As_ppm~Depth, data=USGS_analyze) #tests for homogeneity of variance, p < 0.001

#Does log transforming help meet assumptions for ANOVA?
bartlett.test(As_log~Depth, data=USGS_analyze)#no, p  < 0.001

#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality - this has a built in post hoc test using the criterium Fisher's least significant difference
As_byDepth <- kruskal(USGS_analyze$As_ppm, USGS_analyze$Depth, group=TRUE, p.adj = "bonferroni")
As_byDepth #Chi-sq = 5.996, Df = 2, p = 0.04986
#mid                531.0915      a
#surface            472.3721      a
#deep               463.8210      a


#Getting count by depth for annotating
As_count %>% 
  group_by(Depth) %>%
  summarise(no_rows = length(Depth))

#plot by depth
As_depth <- ggplot(USGS_analyze, aes(depth, As_log, fill=depth)) + 
   geom_boxplot() +
  xlab("source") +
  ylab("ln [As (mg/kg)]") +
  theme_classic()+
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#adds colors and labels
As_depth + 
  scale_fill_manual(values=c("#fee0d2", "#fc9272", "#de2d26"))+
  annotate("text", x = 1:3, y = -0.5, label = c("727", "153", "81"))+
  ylim(-0.5,10) + 
  annotate("text", x = 1:3, y = 10, label = c("a", "a", "a"))

```

Chromium: 
```{r}
#get summary statistics for Cr concentrations
summary(USGS_analyze$Cr_ppm, na.rm=TRUE)
  #Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
    #1.00   33.30   56.00   59.24   78.00  350.00      79  
sd(USGS_analyze$Cr_ppm, na.rm=TRUE)
#stavard deviation is 38.04351

#Checking distribution for stats
shapiro.test(USGS_analyze$Cr_ppm) #tests for normality, p < 0.001
bartlett.test(Cr_ppm~Source, data=USGS_analyze) #tests for homogeneity of variance, p < 0.001

#Does log transforming help meet assumptions for ANOVA?
shapiro.test(USGS_analyze$Cr_log) #no, p < 0.001
bartlett.test(Cr_log~Source, data=USGS_analyze)#no, p  < 0.001

#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality
Cr_bySource <- kruskal(USGS_analyze$Cr_ppm, USGS_analyze$Source, group=TRUE, p.adj = "bonferroni")
Cr_bySource #Chi-sq = 36.008, Df = 4, p < 0.001
#mine-impacted                       863.7895      a
#forested                            585.8810      b
#glacial deposit                     539.7711     bc
#sedge/moss/shrub wetland            535.5665     bc
#tundra                              498.6956      c

#Counts for annotation
Cr_count <- select(USGS_analyze, Source, upper.cm, Depth, Cr_ppm)
Cr_count<- filter(Cr_count, !is.na(Cr_ppm))
Cr_count %>% 
  group_by(Source) %>%
  summarise(no_rows = length(Source))

##Chromium by source boxplot
Cr_plot <- ggplot(USGS_analyze, aes(Source, Cr_log, fill=Source)) + 
  geom_boxplot() +
  xlab("source") +
  ylab("ln [Cr (mg/kg)]") +
  theme_classic()+
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#adds coded colors, number of obs., and post-hoc groups
Cr_plot + 
  scale_fill_manual(values=c("#80b1d3", "#b3de69", "#fdb462", "#fb8072", "#bebada")) + 
  annotate("text", x = 1:5, y = -0.5, label = c("83", "273", "524", "19", "173")) + 
  ylim(-0.5,10) + 
  annotate("text", x = 1:5, y = 10, label = c("bc", "b", "c", "a", "bc"))

#Checking distribution for stats on depth
bartlett.test(Cr_ppm~Depth, data=USGS_analyze) #tests for homogeneity of variance, p = 0.0003218

#Does log transforming help meet assumptions for ANOVA?
bartlett.test(Cr_log~Depth, data=USGS_analyze)#no, p  < 0.001

#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality - this has a built in post hoc test using the criterium Fisher's least significant difference
Cr_byDepth <- kruskal(USGS_analyze$Cr_ppm, USGS_analyze$Depth, group=TRUE, p.adj = "bonferroni")
Cr_byDepth #Chi-sq = 47.137, Df = 2, p <0.001
#mid                632.9935      a
#surface            537.5304      b
#deep               338.6646      c


#getting count for Depth
Cr_count %>% 
 group_by(Depth) %>%
  summarise(no_rows = length(Depth))

#plot by depth
Cr_depth <- ggplot(USGS_analyze, aes(depth, Cr_log, fill=depth)) + 
 geom_boxplot() +
  xlab("source") +
  ylab("ln [Cr (mg/kg)]") +
  theme_classic()+
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

Cr_depth + 
  scale_fill_manual(values=c("#fee6ce", "#fdae6b", "#e6550d")) +
  annotate("text", x = 1:3, y = -0.5, label = c("840", "153", "79")) +
  ylim(-0.5,10) + 
  annotate("text", x = 1:3, y = 10, label = c("b", "a", "c"))


```

Mercury: 
```{r}
#get summary statistics for Hg concentrations
summary(USGS_analyze$Hg_ppm, na.rm=TRUE)
   #Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   # 0.01    0.04    0.06   30.44    0.09 6090.00     351  
sd(USGS_analyze$Hg_ppm, na.rm=TRUE)
#sd is 0.058

#Checking distribution
shapiro.test(USGS_analyze$Hg_ppm) #tests for normality, p < 0.001
bartlett.test(Hg_ppm~Source, data=USGS_analyze) #tests for homogeneity of variance, p < 0.001

#Does log transforming help meet assumptions for ANOVA?
shapiro.test(USGS_analyze$Hg_log) #no, p < 0.001
bartlett.test(Hg_log~Source, data=USGS_analyze)#no, p  < 0.001


#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality
Hg_bySource <- kruskal(USGS_analyze$Hg_ppm, USGS_analyze$Source, group=TRUE, p.adj = "bonferroni")
Hg_bySource #Chi-sq = 72.28, Df = 4, p < 0.001
#mine-impacted                       791.3333      a
#glacial deposit                     489.8313      b
#tundra                              403.1229      c
#sedge/moss/shrub wetland            398.0912      c
#forested                            297.4919      d

#Counts for annotation
Hg_count <- select(USGS_analyze, Source, Depth, upper.cm, Hg_ppm)
Hg_count<- filter(Hg_count, !is.na(Hg_ppm))
Hg_count %>% 
  group_by(Source) %>%
  summarise(no_rows = length(Source))


##Mercury by source boxplot
Hg_plot <- ggplot(USGS_analyze, aes(Source, Hg_log, fill=Source)) + 
  geom_boxplot() +
  xlab("source") +
  ylab("ln [Hg (mg/kg)]") +
  theme_classic()+
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#adds correct colors, number of obs., and post-hoc groups
Hg_plot + 
  scale_fill_manual(values=c("#80b1d3", "#b3de69", "#fdb462", "#fb8072", "#bebada")) + 
  annotate("text", x = 1:5, y = -6, label = c("83", "124", "411", "12", "170")) + 
  ylim(-6,10) + 
  annotate("text", x = 1:5, y = 10, label = c("b", "d", "c", "a", "c")) 

#Checking distribution for stats for depth
bartlett.test(Hg_ppm~Depth, data=USGS_analyze) #tests for homogeneity of variance, p < 0.001

#Does log transforming help meet assumptions for ANOVA?
bartlett.test(Hg_log~Depth, data=USGS_analyze)#no, p  < 0.001

#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality - this has a built in post hoc test using the criterium Fisher's least significant difference
Hg_byDepth <- kruskal(USGS_analyze$Hg_ppm, USGS_analyze$Depth, group=TRUE, p.adj = "bonferroni")
Hg_byDepth #Chi-sq = 6.68, Df = 2, p = 0.0354
#deep               470.5538      a
#surface            395.6481      b
#mid                388.7917      b

#getting depth counts
Hg_count %>% 
  group_by(Depth) %>%
  summarise(no_rows = length(Depth))

#plot by depth
Hg_depth <- ggplot(USGS_analyze, aes(depth, Hg_log, fill=depth)) + 
 geom_boxplot() +
xlab("source") +
  ylab("ln [Hg (mg/kg)]") +
  theme_classic()+
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#adds correct colors and number of obs. 
Hg_depth + 
  scale_fill_manual(values=c("#e5f5e0", "#a1d99b", "#31a354"))+
  annotate("text", x = 1:3, y = -6, label = c("591", "144", "65")) +
  ylim(-6,10) + 
  annotate("text", x = 1:3, y = 10, label = c("b", "b", "a")) 

```

Nickel:
```{r}
#get summary statistics for Ni concentrations
summary(USGS_analyze$Ni_ppm, na.rm=TRUE)
  #Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
    #2.00   19.00   28.30   44.55   41.60  702.00      78
sd(USGS_analyze$Ni_ppm, na.rm=TRUE)
#sd is 60.97551

#Checking distribution
shapiro.test(USGS_analyze$Ni_ppm) #tests for normality, p < 0.001
bartlett.test(Ni_ppm~Source, data=USGS_analyze) #tests for homogeneity of variance, p < 0.0002

#Does log transforming help meet assumptions for ANOVA?
shapiro.test(USGS_analyze$Ni_log) #no, p < 0.001
bartlett.test(Ni_log~Source, data=USGS_analyze)#no, p  < 0.001

#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality
Ni_bySource <- kruskal(USGS_analyze$Ni_ppm, USGS_analyze$Source, group=TRUE, p.adj = "bonferroni")
Ni_bySource #Chi-sq = 140.26, Df = 4, p < 0.001
#mine-impacted                       816.7895      a
#forested                            638.6410      a
#tundra                              553.0659      b
#sedge/moss/shrub wetland            435.2275      c
#glacial deposit                     240.6265      d

##gets counts for annotation
Ni_count <- select(USGS_analyze, Source, Depth, upper.cm, Ni_ppm)
Ni_count<- filter(Ni_count, !is.na(Ni_ppm))
Ni_count %>% 
  group_by(Source) %>%
  summarise(no_rows = length(Source))


##Nickel by source boxplot
Ni_plot <- ggplot(USGS_analyze, aes(Source, Ni_log, fill=Source)) + 
  geom_boxplot() +
  xlab("source") +
  ylab("ln [Ni (mg/kg)]") +
  theme_classic()+ 
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#adds correct colors, number of obs., lines for WHO limit/US avg, and post-hoc groups 
Ni_plot + 
  scale_fill_manual(values=c("#80b1d3", "#b3de69", "#fdb462", "#fb8072", "#bebada")) + 
  annotate("text", x = 1:5, y = -0.5, label = c("83", "273", "531", "19", "167")) +
  ylim(-0.5,10) + 
  annotate("text", x = 1:5, y = 10, label = c("d", "a", "b", "a", "c"))

#Checking distribution for stats for depth
bartlett.test(Ni_ppm~Depth, data=USGS_analyze) #tests for homogeneity of variance, p < 0.001

#Does log transforming help meet assumptions for ANOVA?
bartlett.test(Ni_log~Depth, data=USGS_analyze) #no, p < 0.001

#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality - this has a built in post hoc test using the criterium Fisher's least significant difference
Ni_byDepth <- kruskal(USGS_analyze$Ni_ppm, USGS_analyze$Depth, group=TRUE, p.adj = "bonferroni")
Ni_byDepth #Chi-sq = 63.37, Df = 2, p < 0.001
#surface            577.1455      a
#mid                402.3758      b
#deep               384.9529      b

#getting depth count
Ni_count %>% 
  group_by(Depth) %>%
  summarise(no_rows = length(Depth))

#depth plot
Ni_depth <- ggplot(USGS_analyze, aes(depth, Ni_log, fill=depth)) + 
  geom_boxplot() +
  xlab("depth") +
  ylab("ln [Ni (mg/kg)]") +
  theme_classic()+ 
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#adds correct colors and number of obs. 
Ni_depth + 
  scale_fill_manual(values=c("#deebf7", "#9ecae1","#3182bd")) +
  annotate("text", x = 1:3, y = -0.5, label = c("835", "153", "85")) +
  ylim(-0.5,10) + 
  annotate("text", x = 1:3, y = 10, label = c("a", "b", "b"))
```

Lead:
```{r}
#get summary statistics for Pb concentrations
summary(USGS_analyze$Pb_ppm, na.rm=TRUE)
    #Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   #2.00    9.00   13.00   21.06   19.00  720.00      70 
sd(USGS_analyze$Pb_ppm, na.rm=TRUE)
#stavard deviation is 41.14311

#Checking distrubition
shapiro.test(USGS_analyze$Pb_ppm) #tests for normality, p < 0.001
bartlett.test(Pb_ppm~Source, data=USGS_analyze) #tests for homogeneity of variance, p < 0.001

#Does log transforming help meet assumptions for ANOVA?
shapiro.test(USGS_analyze$Pb_log) #no, p < 0.001
bartlett.test(Pb_log~Source, data=USGS_analyze)#no, p  < 0.001

#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality
Pb_bySource <- kruskal(USGS_analyze$Pb_ppm, USGS_analyze$Source, group=TRUE, p.adj = "bonferroni")
Pb_bySource #Chi-sq = 119.91, Df = 4, p < 0.001
#mine-impacted                       885.1579      a
#forested                            580.6905      b
#glacial deposit                     575.9699      b
#tundra                              573.2651      b
#sedge/moss/shrub wetland            317.3114      c

##getting counts for each source type to annotate figure 
Pb_count <- select(USGS_analyze, Source, Depth, upper.cm, Pb_ppm, Pb_log)
Pb_count<- filter(Pb_count, !is.na(Pb_ppm))
Pb_count %>% 
  group_by(Source) %>%
  summarise(no_rows = length(Source))

##Lead by source boxplot
Pb_plot <- ggplot(USGS_analyze, aes(Source, Pb_log, fill=Source)) + 
   geom_boxplot() +
  xlab("source") +
  ylab("ln [Pb (mg/kg)]") +
  theme_classic()+
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

#adds correct colors, number of obs., lines for WHO limit/US avg, and post-hoc groups 
Pb_plot + 
  scale_fill_manual(values=c("#80b1d3", "#b3de69", "#fdb462", "#fb8072", "#bebada"))+
  annotate("text", x = 1:5, y = -0.5, label = c("83", "231", "581", "19", "167")) +
  ylim(-0.5,10) +
  annotate("text", x = 1:5, y = 10, label = c("b", "b", "b", "a", "c"))

#Checking distribution for stats for depth
bartlett.test(Pb_ppm~Depth, data=USGS_analyze) #tests for homogeneity of variance, p < 0.001

#Does log transforming help meet assumptions for ANOVA?
bartlett.test(Pb_log~Depth, data=USGS_analyze) #no, p < 0.001

#using Kruskal-Wallis test in lieu of an ANOVA because of the lack of normality - this has a built in post hoc test using the criterium Fisher's least significant difference
Pb_byDepth <- kruskal(USGS_analyze$Pb_ppm, USGS_analyze$Depth, group=TRUE, p.adj = "bonferroni")
Pb_byDepth #Chi-sq = 11.385, Df = 2, p = 0.0034
#mid                620.0822      a
#surface            528.1214      b
#deep               527.4588      b

#depth counts
Pb_count %>% 
  group_by(Depth) %>%
  summarise(no_rows = length(Depth))

#By depth
Pb_depth <- ggplot(USGS_analyze, aes(depth, Pb_log, fill=depth)) + 
   geom_boxplot() +
  xlab("source") +
  ylab("ln [Pb (mg/kg)]") +
  theme_classic()+
   theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

Pb_depth + 
  scale_fill_manual(values=c("#efedf5", "#bcbddc","#756bb1")) +
  annotate("text", x = 1:3, y = -0.5, label = c("844","152","85"))  +
  ylim(-0.5,10) +
  annotate("text", x = 1:3, y = 10, label = c("b", "a", "b"))

```

Create dataframe and plots of proportion of samples over US averages, and greater than averages by 2x, 10x, and 1 standard deviation.Comparison values from A Horizon soils from the USGS, Smith et al., 2013, Geochem. and Mineralogical Data for Soils of the Conterminous US, Data Series 801
```{r}
df4 <- data.frame(metal= rep(c("As", "Cr", "Hg", "Ni", "Pb"), 4),
                  limit=rep(c("above average","2x average","10x average", "plus 1 SD"),each = 5),
                  percent = c(mean((USGS_analyze$As_ppm > 6.6)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Cr_ppm > 37)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Hg_ppm > 0.04)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Ni_ppm > 18.5)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Pb_ppm > 22.2)*100, na.rm=TRUE),
                              mean((USGS_analyze$As_ppm > 13.2)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Cr_ppm > 74)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Hg_ppm > 0.08)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Ni_ppm > 37)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Pb_ppm > 44.4)*100, na.rm=TRUE),
                              mean((USGS_analyze$As_ppm > 66)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Cr_ppm > 370)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Hg_ppm > 0.4)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Ni_ppm > 185)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Pb_ppm > 222)*100, na.rm=TRUE),
                              mean((USGS_analyze$As_ppm > 26.2)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Cr_ppm > 126)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Hg_ppm > 0.21)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Ni_ppm > 72.9)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Pb_ppm > 68.8)*100, na.rm=TRUE)))
#Barchart in figure 3
ggplot(data=df4, aes(x=metal, y=percent, fill=limit)) +
  geom_bar(stat="identity", position=position_dodge()) + ylim(0,100) +
  geom_text(aes(label=round(percent, 1)), vjust=-0.2, color="black",
            position = position_dodge(0.9), size = 3.25, fontface = "bold") +
  scale_fill_manual(values=c("#984ea3", "#377eb8", "#4daf4a", "#e41a1c"))+
  ylab("Percent of sample locations over value")+
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_text(face="bold"),
        axis.text.x = element_text(size = 10, face = "bold"),
        axis.text.y = element_text(size = 10, face = "bold"),
        legend.text=element_text(face="bold"),
        legend.title=element_blank())

#Making a dataframe with no mine-impacted soil to compare % levels
no_mine <- filter(USGS_analyze, Source != "mine-impacted")

#DF to compare % with and without mine-impacted sites included
df5 <- data.frame(metal= rep(c("As", "Cr", "Hg", "Ni", "Pb"), 2),
                  mines= rep(c("with mines", "without"), each = 5),
                  percent = c(mean((USGS_analyze$As_ppm > 26.2)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Cr_ppm > 126)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Hg_ppm > 0.21)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Ni_ppm > 72.9)*100, na.rm=TRUE), 
                              mean((USGS_analyze$Pb_ppm > 68.8)*100, na.rm=TRUE), 
                              mean((no_mine$As_ppm > 26.2)*100, na.rm=TRUE), 
                              mean((no_mine$Cr_ppm > 126)*100, na.rm=TRUE), 
                              mean((no_mine$Hg_ppm > 0.21)*100, na.rm=TRUE), 
                              mean((no_mine$Ni_ppm > 72.9)*100, na.rm=TRUE), 
                              mean((no_mine$Pb_ppm > 68.8)*100, na.rm=TRUE)))
#barchart for figure S2
ggplot(data=df5, aes(x=metal, y=percent, fill=mines)) +
  scale_fill_manual(values=c("#bdbdbd", "#737373"))+
  geom_text(aes(label=round(percent, 2)), vjust=-0.2, color="black",
            position = position_dodge(0.9), size = 3.25, fontface = "bold") +
  ylab("percent of samples with elevated concentration")+
  geom_bar(stat="identity", position=position_dodge()) + ylim(0,25) + theme_classic()+
  theme(axis.title.x=element_blank(),
        axis.title.y=element_text(face="bold"),
        axis.text.x = element_text(size = 10, face = "bold"),
        axis.text.y = element_text(size = 10, face = "bold"),
        legend.text=element_text(face="bold"),
        legend.title=element_blank())
```
